The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-06 manuscript/manuscript.Rnw
fix cgb table link

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:chorionic_gonadotropin}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2}),
insulin growth factor 2 (\textit{IGF2}) and one of its binding
proteins \textit{IGFBP3}, and the hormone adrenomedullin
(\textit{ADM}).

>>>>

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:cgb_evolution_tree}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2}),
insulin growth factor 2 (\textit{IGF2}) and one of its binding
proteins \textit{IGFBP3}, and the hormone adrenomedullin
(\textit{ADM}).
